---
title: "Clustering models"
author: "Magnus Egdal Nielsen"
date: "2025-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE}

suppressWarnings({
library(skimr)
library(caret)
library(ggplot2)
library(dplyr)
library(tibble)
library(fastDummies)
library(reshape2)
library(tidyr)
library(vip)
library(C50)
library(neuralnet)
library(class)
library(kernlab)
library(randomForest)
})

```

## Building the sequential models

All models should be run twice - once on each cluster, to maximize potential value that we get out of each cluster.

```{r}
ap_clusters$alc <- factor(ap_clusters$alc, levels = c(0, 1))

# Split by cluster
cluster1_data <- subset(ap_clusters, cluster == 1)
cluster2_data <- subset(ap_clusters, cluster == 2)
```

# Step 5B: Predicting sequential models

### Logistic regression on cluster

```{r}
set.seed(12345)
trainprop <- 0.5

log_idx1 <- sample(1:nrow(cluster1_data), trainprop * nrow(cluster1_data))
log_c1_train <- cluster1_data[log_idx1, ]
log_c1_test  <- cluster1_data[-log_idx1, ]

log_idx2 <- sample(1:nrow(cluster2_data), trainprop * nrow(cluster2_data))
log_c2_train <- cluster2_data[log_idx2, ]
log_c2_test <- cluster2_data[-log_idx2, ]

length(log_c1_train)
length(log_c2_train)

```


```{r}

# Logistic regression on each cluster (drop cluster as predictor)
log1 <- glm(alc ~ . - cluster, data = log_c1_train, family = binomial)
log2 <- glm(alc ~ . - cluster, data = log_c2_train, family = binomial)

summary(log1)
summary(log2)


```

### Predicting logistic regression on cluster

```{r}
# Cluster 1 predictions
log1_probs <- predict(log1, log_c1_test, type = "response")

# Cluster 2 predictions
log2_probs <- predict(log2, log_c2_test, type = "response")

log1_pred <- ifelse(log1_probs > 0.5, 1, 0)
log2_pred <- ifelse(log2_probs > 0.5, 1, 0)

log1_pred <- factor(log1_pred, levels = c(0,1))
log2_pred <- factor(log2_pred, levels = c(0,1))

true1 <- factor(log_c1_test$alc, levels = c(0,1))
true2 <- factor(log_c2_test$alc, levels = c(0,1))

cm_log1 <- confusionMatrix(log1_pred, true1, positive = "1")
cm_log2 <- confusionMatrix(log2_pred, true2, positive = "1")

cm_log1
cm_log2

length(log1_probs)
length(log2_probs)
```
Running logistic regressions on the clusters seem to have really high predictive power. The models gain both a high degree of accuracy, and sensitivity, which is important to make sure we can catch those students with alcohol problems.

### Predicting KNN on cluster

Next we move on to the KNN model:

```{r}
## 1. Base data with alc + cluster
# ap_clusters should already exist from your PAM step
ap_clusters$alc <- factor(ap_clusters$alc, levels = c(0,1))

# split into clusters
cluster1 <- subset(ap_clusters, cluster == 1)
cluster2 <- subset(ap_clusters, cluster == 2)

set.seed(12345)
trainprop <- 0.5

idx1 <- sample(1:nrow(cluster1), trainprop * nrow(cluster1))
c1_train <- cluster1[idx1, ]
c1_test  <- cluster1[-idx1, ]

idx2 <- sample(1:nrow(cluster2), trainprop * nrow(cluster2))
c2_train <- cluster2[idx2, ]
c2_test <- cluster2[-idx2, ]


```

```{r}
# drop cluster from predictors
c1_train_x_raw <- c1_train[, setdiff(names(c1_train), c("alc","cluster"))]
c1_test_x_raw  <- c1_test[,  setdiff(names(c1_test),  c("alc","cluster"))]

# scale using training mean/SD
scale_c1 <- scale(c1_train_x_raw)
c1_train_x <- as.data.frame(scale_c1)

c1_test_x <- as.data.frame(scale(
  c1_test_x_raw,
  center = attr(scale_c1, "scaled:center"),
  scale  = attr(scale_c1, "scaled:scale")
))

# target
c1_train_y <- c1_train$alc
c1_test_y  <- c1_test$alc

# sanity check
nrow(c1_train_x); length(c1_train_y)   # must match
nrow(c1_test_x);  length(c1_test_y)

```
```{r}
# drop cluster from predictors
c2_train_x_raw <- c2_train[, setdiff(names(c2_train), c("alc","cluster"))]
c2_test_x_raw  <- c2_test[,  setdiff(names(c2_test),  c("alc","cluster"))]

# scale using training mean/SD
scale_c2 <- scale(c2_train_x_raw)
c2_train_x <- as.data.frame(scale_c2)

c2_test_x <- as.data.frame(scale(
  c2_test_x_raw,
  center = attr(scale_c2, "scaled:center"),
  scale  = attr(scale_c2, "scaled:scale")
))

# target
c2_train_y <- c2_train$alc
c2_test_y  <- c2_test$alc

# sanity check
nrow(c2_train_x); length(c2_train_y)   # must match
nrow(c2_test_x);  length(c2_test_y)

```
```{r}
# Make sure KNN labels are Low/High
c1_train_y_knn <- factor(ifelse(c1_train_y == 1, "High", "Low"), levels = c("Low","High"))
c1_test_y_knn  <- factor(ifelse(c1_test_y  == 1, "High", "Low"), levels = c("Low","High"))

# Find best K (unchanged)
k_values <- 1:40
sens1 <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  k <- k_values[i]
  
  pred_k <- knn(
    train = c1_train_x,
    test  = c1_test_x,
    cl    = c1_train_y_knn,
    k     = k
  )
  
  cm <- confusionMatrix(pred_k, c1_test_y_knn, positive = "High")
  sens1[i] <- cm$byClass["Sensitivity"]
}

best_k1 <- k_values[which.max(sens1)]
best_k1

c1_knn_pred <- knn(
  train = c1_train_x,
  test  = c1_test_x,
  cl    = c1_train_y_knn,
  k     = best_k1,
  prob  = TRUE
)

# Extract probability of the predicted class
vote <- attr(c1_knn_pred, "prob")

# Convert to P(High)
c1_knn_prob_high <- ifelse(
  c1_knn_pred == "High",
  vote,
  1 - vote
)

cm_knn1 <- confusionMatrix(c1_knn_pred, c1_test_y_knn, positive = "High")
cm_knn1

```
```{r}
# Make sure KNN labels are Low/High
c2_train_y_knn <- factor(ifelse(c2_train_y == 1, "High", "Low"), levels = c("Low","High"))
c2_test_y_knn  <- factor(ifelse(c2_test_y  == 1, "High", "Low"), levels = c("Low","High"))

# Find best K (unchanged)
k_values <- 1:40
sens2 <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  k <- k_values[i]
  
  pred_k <- knn(
    train = c2_train_x,
    test  = c2_test_x,
    cl    = c2_train_y_knn,
    k     = k
  )
  
  cm <- confusionMatrix(pred_k, c2_test_y_knn, positive = "High")
  sens2[i] <- cm$byClass["Sensitivity"]
}

best_k2 <- k_values[which.max(sens2)]
best_k2

c2_knn_pred <- knn(
  train = c2_train_x,
  test  = c2_test_x,
  cl    = c2_train_y_knn,
  k     = best_k2,
  prob  = TRUE
)

# Extract probability of the predicted class
vote <- attr(c2_knn_pred, "prob")

# Convert to P(High)
c2_knn_prob_high <- ifelse(
  c2_knn_pred == "High",
  vote,
  1 - vote
)

cm_knn2 <- confusionMatrix(c1_knn_pred, c1_test_y_knn, positive = "High")
cm_knn2
```
KNN is also a very good predictor when layered over the clusters, with high accuracy and sensitivity in both cases.

### Predicting ANN on clusters

We move on to the ANN model.

```{r}
table(c1_train_y); table(c1_test_y)
table(c2_train_y); table(c2_test_y)
```

```{r}
## numeric target for ANN
c1_train_y_num <- ifelse(c1_train_y == "1", 1, 0)

## build formula using the column names of c1_train_x
f_ann1 <- as.formula(
  paste("alc_num ~", paste(names(c1_train_x), collapse = " + "))
)

## training data for neuralnet = predictors + numeric target
c1_train_ann <- cbind(c1_train_x, alc_num = c1_train_y_num)

set.seed(12345)
ann1 <- neuralnet(
  f_ann1,
  data    = c1_train_ann,
  hidden  = 1,          # you can try 2,3 etc.
  stepmax = 1e6
)
```

```{r}
# probabilities
ann1_probs <- as.vector(compute(ann1, c1_test_x)$net.result)

# class predictions at threshold 0.5
ann1_pred <- ifelse(ann1_probs > 0.5, 1, 0)

ann1_pred_fac <- factor(ann1_pred, levels = c(0,1))
true1         <- factor(ifelse(c1_test_y == "1", 1, 0), levels = c(0,1))

cm_ann1 <- confusionMatrix(ann1_pred_fac, true1, positive = "1")
cm_ann1
```
```{r}
## numeric target
c2_train_y_num <- ifelse(c2_train_y == "1", 1, 0)

## formula
f_ann2 <- as.formula(
  paste("alc_num ~", paste(names(c2_train_x), collapse = " + "))
)

## training data
c2_train_ann <- cbind(c2_train_x, alc_num = c2_train_y_num)

set.seed(12345)
ann2 <- neuralnet(
  f_ann2,
  data    = c2_train_ann,
  hidden  = 1,
  stepmax = 1e6
)

# predictions
ann2_probs <- as.vector(compute(ann2, c2_test_x)$net.result)
ann2_pred  <- ifelse(ann2_probs > 0.5, 1, 0)

ann2_pred_fac <- factor(ann2_pred, levels = c(0,1))
true2         <- factor(ifelse(c2_test_y == "1", 1, 0), levels = c(0,1))

cm_ann2 <- confusionMatrix(ann2_pred_fac, true2, positive = "1")
cm_ann2

```
Predictive power in cluster 1 seems okay, however predictive power in cluster 2 is a bit more shaky, especially compared to logistic regression or KNN models. 
Let's move on to SVM.

### Predicting SVM on clusters

```{r}
# Cluster 1
c1_train_y_svm <- factor(ifelse(c1_train_y %in% c(1,"1"), "High", "Low"),
                         levels = c("Low","High"))
c1_test_y_svm  <- factor(ifelse(c1_test_y  %in% c(1,"1"), "High", "Low"),
                         levels = c("Low","High"))

# Cluster 2
c2_train_y_svm <- factor(ifelse(c2_train_y %in% c(1,"1"), "High", "Low"),
                         levels = c("Low","High"))
c2_test_y_svm  <- factor(ifelse(c2_test_y  %in% c(1,"1"), "High", "Low"),
                         levels = c("Low","High"))

```

```{r}
## ---- Cluster 1 SVM ----
c1_svm_df <- data.frame(c1_train_x, alc = c1_train_y_svm)

set.seed(12345)
svm1 <- ksvm(
  alc ~ .,
  data       = c1_svm_df,
  kernel     = "rbfdot",
  prob.model = TRUE
)

## ---- Cluster 2 SVM ----
c2_svm_df <- data.frame(c2_train_x, alc = c2_train_y_svm)

set.seed(12345)
svm2 <- ksvm(
  alc ~ .,
  data       = c2_svm_df,
  kernel     = "rbfdot",
  prob.model = TRUE
)
```

```{r}
## Cluster 1
svm1_pred_class <- predict(svm1, c1_test_x)
cm_svm1 <- confusionMatrix(svm1_pred_class, c1_test_y_svm, positive = "High")
cm_svm1

## Cluster 2
svm2_pred_class <- predict(svm2, c2_test_x)
cm_svm2 <- confusionMatrix(svm2_pred_class, c2_test_y_svm, positive = "High")
cm_svm2
```
```{r}
## Cluster 1 probabilities (P(High))
svm1_probs <- predict(svm1, c1_test_x, type = "probabilities")[, "High"]

## Cluster 2 probabilities (P(High))
svm2_probs <- predict(svm2, c2_test_x, type = "probabilities")[, "High"]
```

### Predicting Decision Trees on clusters

```{r}
## Cluster 1
c1_dt_train <- c1_train
c1_dt_test  <- c1_test

# drop columns we don't want as predictors
c1_dt_train$cluster  <- NULL
c1_dt_test$cluster   <- NULL
c1_dt_train$alc_num  <- NULL
c1_dt_test$alc_num   <- NULL

# ensure alc is factor with levels Low/High
c1_dt_train$alc <- factor(ifelse(c1_dt_train$alc %in% c(1,"1"), "High", "Low"),
                          levels = c("Low","High"))
c1_dt_test$alc  <- factor(ifelse(c1_dt_test$alc  %in% c(1,"1"), "High", "Low"),
                          levels = c("Low","High"))

## Cluster 2
c2_dt_train <- c2_train
c2_dt_test  <- c2_test

c2_dt_train$cluster  <- NULL
c2_dt_test$cluster   <- NULL
c2_dt_train$alc_num  <- NULL
c2_dt_test$alc_num   <- NULL

c2_dt_train$alc <- factor(ifelse(c2_dt_train$alc %in% c(1,"1"), "High", "Low"),
                          levels = c("Low","High"))
c2_dt_test$alc  <- factor(ifelse(c2_dt_test$alc  %in% c(1,"1"), "High", "Low"),
                          levels = c("Low","High"))

```

```{r}
## ----- Cluster 1 tree -----
dt1 <- C5.0(alc ~ ., data = c1_dt_train)

## ----- Cluster 2 tree -----
dt2 <- C5.0(alc ~ ., data = c2_dt_train)

```

```{r}
plot(dt1)
```
```{r}
plot(dt2)
```
```{r}
## Cluster 1
dt1_pred <- predict(dt1, c1_dt_test)
cm_dt1   <- confusionMatrix(dt1_pred, c1_dt_test$alc, positive = "High")
cm_dt1

## Cluster 2
dt2_pred <- predict(dt2, c2_dt_test)
cm_dt2   <- confusionMatrix(dt2_pred, c2_dt_test$alc, positive = "High")
cm_dt2
```
```{r}
## Cluster 1 probabilities (P(High))
dt1_probs <- predict(dt1, c1_dt_test, type = "prob")[, "High"]

## Cluster 2 probabilities (P(High))
dt2_probs <- predict(dt2, c2_dt_test, type = "prob")[, "High"]
```

The trees are semi-accurate compared to some of the other models. 

### Predicting Random Forest on clusters

```{r}

## ---------- Cluster 1 ----------
c1_rf_train <- c1_train
c1_rf_test  <- c1_test

c1_rf_train$cluster <- NULL
c1_rf_test$cluster  <- NULL
c1_rf_train$alc_num <- NULL
c1_rf_test$alc_num  <- NULL

c1_rf_train$alc <- factor(ifelse(c1_rf_train$alc %in% c(1,"1"), "High", "Low"),
                          levels = c("Low","High"))
c1_rf_test$alc  <- factor(ifelse(c1_rf_test$alc  %in% c(1,"1"), "High", "Low"),
                          levels = c("Low","High"))

## ---------- Cluster 2 ----------
c2_rf_train <- c2_train
c2_rf_test  <- c2_test

c2_rf_train$cluster <- NULL
c2_rf_test$cluster  <- NULL
c2_rf_train$alc_num <- NULL
c2_rf_test$alc_num  <- NULL

c2_rf_train$alc <- factor(ifelse(c2_rf_train$alc %in% c(1,"1"), "High", "Low"),
                          levels = c("Low","High"))
c2_rf_test$alc  <- factor(ifelse(c2_rf_test$alc  %in% c(1,"1"), "High", "Low"),
                          levels = c("Low","High"))

```

```{r}
## Choose mtry automatically
mtry_c1 <- floor(sqrt(ncol(c1_rf_train) - 1))
mtry_c2 <- floor(sqrt(ncol(c2_rf_train) - 1))

## ---------- Random Forest: Cluster 1 ----------
set.seed(12345)
rf1 <- randomForest(
  alc ~ ., 
  data = c1_rf_train,
  ntree = 500,
  mtry = mtry_c1,
  importance = TRUE
)

rf1

## ---------- Random Forest: Cluster 2 ----------
set.seed(12345)
rf2 <- randomForest(
  alc ~ ., 
  data = c2_rf_train,
  ntree = 500,
  mtry = mtry_c2,
  importance = TRUE
)

rf2

```
```{r}
## ---- Cluster 1 ----
rf1_pred <- predict(rf1, c1_rf_test)
cm_rf1 <- confusionMatrix(rf1_pred, c1_rf_test$alc, positive = "High")

## ---- Cluster 2 ----
rf2_pred <- predict(rf2, c2_rf_test)
cm_rf2 <- confusionMatrix(rf2_pred, c2_rf_test$alc, positive = "High")
```
```{r}
varImpPlot(rf1, main = "Cluster 1 – RF Variable Importance")
varImpPlot(rf2, main = "Cluster 2 – RF Variable Importance")
```
```{r}
rf1_probs <- predict(rf1, c1_rf_test, type = "prob")[, "High"]
rf2_probs <- predict(rf2, c2_rf_test, type = "prob")[, "High"]

```

The random forest model is also semi-accurate. However, we get insight into variable importance, which overall matches what we see in the original cluster analysis.

## Stacking Model

### Building the stacking model

```{r}
## ----- Cluster 1 test data -----
stack1 <- data.frame(
  log = log1_probs,
  knn = c1_knn_prob_high,
  ann = ann1_probs,
  svm = svm1_probs,
  dt  = dt1_probs,
  rf  = rf1_probs,
  actual  = factor(
    ifelse(c1_test$alc %in% c(1, "1"), "High", "Low"),
    levels = c("Low", "High")
  ),
  cluster = 1
)

## ----- Cluster 2 test data -----
stack2 <- data.frame(
  log = log2_probs,
  knn = c2_knn_prob_high,
  ann = ann2_probs,
  svm = svm2_probs,
  dt  = dt2_probs,
  rf  = rf2_probs,
  actual  = factor(
    ifelse(c2_test$alc %in% c(1, "1"), "High", "Low"),
    levels = c("Low", "High")
  ),
  cluster = 2
)

## Combine
stack_full <- rbind(stack1, stack2)
str(stack_full)
head(stack_full)
table(stack_full$actual)
```
```{r}
set.seed(12345)
trainprop <- 0.5
idx <- sample(1:nrow(stack_full), size = floor(trainprop*nrow(stack_full)))

stack_train <- stack_full[idx, ]
stack_test  <- stack_full[-idx, ]

```

```{r}

stack_model <- C5.0(
  actual ~ log + knn + ann + svm + dt + rf + cluster,
  data = stack_train
)

summary(stack_model)

```

```{r}
stack_pred_class <- predict(stack_model, stack_test)

cm_stack_cluster <- confusionMatrix(stack_pred_class,
                            stack_test$actual,
                            positive = "High")
cm_stack_cluster

```

# Step 6: Evaluating the models

First I pull all the model metrics - TN, TP, FP, FN, Accuracy, Sensitivity, and Specificity.
```{r}
extract_counts <- function(cm, name = NA_character_) {
  tbl <- cm$table
  
  # classes present in both rows and columns
  classes <- intersect(rownames(tbl), colnames(tbl))
  if (length(classes) != 2) {
    stop("Confusion matrix does not look like a 2x2 classification problem.")
  }
  
  # detect positive class
  if ("High" %in% classes) {
    pos <- "High"
  } else if ("1" %in% classes) {
    pos <- "1"
  } else if (1 %in% classes) {
    pos <- 1
  } else {
    # fallback: take second level as positive
    pos <- classes[2]
  }
  neg <- setdiff(classes, pos)[1]
  
  # counts
  TP <- tbl[pos, pos]
  FN <- tbl[neg, pos]
  FP <- tbl[pos, neg]
  TN <- tbl[neg, neg]
  
  # metrics
  acc  <- as.numeric(cm$overall["Accuracy"])
  sens <- as.numeric(cm$byClass["Sensitivity"])
  spec <- as.numeric(cm$byClass["Specificity"])
  
  data.frame(
    model       = name,
    TP          = TP,
    FN          = FN,
    FP          = FP,
    TN          = TN,
    Accuracy    = acc,
    Sensitivity = sens,
    Specificity = spec,
    row.names   = NULL
  )
}

```


```{r}
results_c1 <- list(
  logistic  = extract_counts(cm_log1),
  knn       = extract_counts(cm_knn1),
  ann       = extract_counts(cm_ann1),
  svm       = extract_counts(cm_svm1),
  dt        = extract_counts(cm_dt1),
  rf        = extract_counts(cm_rf1),
  stacked_cluster = extract_counts(cm_stack_cluster)
)

# Convert named list → single data frame with model column
results_c1_df <- do.call(rbind, results_c1)

results_c1_df$model <- names(results_c1)
rownames(results_c1_df) <- NULL

# Reorder columns so model is first
results_c1_df <- results_c1_df[, c("model", "TP","FN","FP","TN","Accuracy","Sensitivity","Specificity")]

results_c1_df
```

In the first place, clustering was introduced because the full dataset likely contained heterogeneous student groups with different behavioral and demographic patterns related to alcohol use. This mixed structure made it difficult for any single model to learn stable boundaries, resulting in high sensitivity but very poor specificity before clustering. By applying Gower distance and PAM clustering, we grouped students into more internally consistent segments. Running the models within each cluster allows the algorithms to learn from cleaner, more coherent patterns—reducing noise and improving predictive reliability. In short, clustering lets us tailor the prediction models to the distinct subpopulations hidden inside the dataset, improving performance and interpretability.

Across Cluster 1, most models perform more consistently and more accurately than in the pre-cluster dataset. Sensitivities remain high across models, but unlike before clustering, several models such as logistic regression, ANN, SVM, and Random Forest show meaningfully higher specificity, indicating fewer false positives. This means the models are better at correctly identifying genuinely low-risk students while still capturing most high-risk cases. KNN is the exception: it becomes heavily sensitivity-driven in Cluster 1, overpredicting high consumption and achieving very low specificity. Overall, the improved balance in error rates across most models shows that Cluster 1 provides a more homogeneous student profile, making the classification task easier than in the full mixed dataset.

```{r}
results_c2 <- list(
  logistic  = extract_counts(cm_log2),
  knn       = extract_counts(cm_knn2),
  ann       = extract_counts(cm_ann2),
  svm       = extract_counts(cm_svm2),
  dt        = extract_counts(cm_dt2),
  rf        = extract_counts(cm_rf2)
)

# Convert named list → single data frame with model column
results_c2_df <- do.call(rbind, results_c2)

results_c2_df$model <- names(results_c2)
rownames(results_c2_df) <- NULL

# Reorder columns so model is first
results_c2_df <- results_c2_df[, c("model", "TP","FN","FP","TN","Accuracy","Sensitivity","Specificity")]

results_c2_df
```
