---
title: "Predicting Tournament Success with Stacked Ensemble"
author: "Caleb Hyun"
date: "November 6, 2025"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(ggplot2)
library(caret)
library(class)
library(neuralnet)
library(kernlab)
library(C50)
library(reshape2)
library(dplyr)
library(knitr)
library(randomForest)
library(e1071)
library(pROC)
library(cluster)

set.seed(12345)
```

## 1 Why do we Care?

Our goal is to identify which college basketball teams are most likely to make a Final Four appearance based on their regular season metrics. Early identification of these "Cinderella" or powerhouse teams allows athletic departments, analysts, and fans to allocate their attention and resources effectively. The "best" model will effectively identify tournament contenders (True Positives) while minimizing false hope for teams that won't make it deep (False Positives).

## 2 Load the Data

```{r}
br <- read.csv("march.csv")

# Turn All String into Factor
br[sapply(br, is.character)] <- lapply(br[sapply(br, is.character)], as.factor)

br <- br %>%
  rename(finals = Final.Four.)
 
br$finals <- ifelse(br$finals == "No", 0, 1)

# Preserve original data with Season and team names for later matching
# We'll filter this the same way as br to keep rows aligned
br_original <- br[, c("Season", "Full.Team.Name")]

summary <- factor(br$finals, levels = c(0, 1), labels = c("No", "Yes"))
y_summary <- br |>
  group_by(finals) |>
  summarise(Count = n()) |>
  mutate(Percent = round(Count / sum(Count) * 100, 1))

kable(y_summary, col.names = c("Finals", "Count", "Percent"))
```

## 3 Clean The Data

Next, we clean the data. We remove variations of redundant data columns such as Short Conference Name and Mapped Conference Name and Mapped ESPN Team Name, as well as specific seasons that were problematic.

```{r}
br <- br[, !sapply(br, is.factor) | (names(br) %in% c("finals", "Active.Coaching.Length"))]

br <- subset(br, Season > 2010) 
br <- subset(br, Season != 2020)

# Filter metadata the same way
br_original <- br_original[br_original$Season > 2010 & br_original$Season != 2020, ]

br$Post.Season.Tournament.Sorting.Index <- NULL

br$Since <- NULL
br$Active.Coaching.Length <- NULL

br <- br[, colSums(is.na(br)) == 0]
# Note: colSums removes columns, not rows, so br_original stays aligned

cols <- grepl("Raw", names(br))
br <- br[, !cols]

cols <- grepl("Rank", names(br))
br <- br[, !cols]
```

Now we create dummy variables and scale the data. We use min-max scaling to normalize the feature range for distance-based algorithms like KNN.

```{r}
br_d <- as.data.frame(model.matrix(~ . -1, data = br))

minmax <- function(x){
  (x - min(x)) / (max(x) - min(x))
}

br_minmax <- as.data.frame(lapply(br_d, minmax))
```

## 4 Cluster Analysis

We use K-means clustering to segment the teams into 4 distinct groups, hoping to isolate different performance profiles. We then filter to keep only Clusters 1 and 3 for our predictive modeling.

```{r}
set.seed(12345)

br_cluster_data <- br_minmax[, names(br_minmax) != "finals"]

k4 <- kmeans(br_cluster_data, centers = 4)

br_minmax$cluster <- k4$cluster

br_filtered <- subset(br_minmax, cluster %in% c(1, 3))
br_filtered$cluster <- factor(br_filtered$cluster)

table(br_filtered$cluster)
```

## 5 Split Train Test (Level 1)

**Data Flow Summary:**
1. **Load and Clean Data**: Load raw data, clean columns, remove problematic seasons, create dummy variables, and apply min-max scaling.
2. **Cluster Analysis**: Perform K-means clustering (k=4) on the scaled data to identify distinct team performance profiles.
3. **Filter Clusters**: Keep only Clusters 1 and 3 for predictive modeling (removing Clusters 2 and 4).
4. **Split Data**: Divide the filtered data into:
   - **Training Set (70%)**: To train Level 1 models and Level 2 model
   - **Test Set (30%)**: To evaluate ALL models fairly on unseen data
5. **Train Level 1 Models**: Build three models (Logistic Regression, Decision Tree, Random Forest) on the training set. KNN and SVM are excluded as they don't handle class imbalance well. We use a cost matrix in Level 2 to handle class imbalance rather than upsampling.
6. **Generate Level 1 Predictions**: 
   - Predict on **training set** → used as features for Level 2 training
   - Predict on **test set** → used for final evaluation
7. **Train Level 2 Model**: Build stacked model using Level 1 predictions from training set, with cost matrix (10:1) to penalize False Negatives.
8. **Evaluate**: Test Level 2 model on test set predictions.

We'll split the **filtered data (Clusters 1 & 3)** into two sets:
1. **Training Set (70%)**: To train Level 1 models and the Level 2 (Stacked) model.
2. **Test Set (30%)**: To evaluate ALL models fairly on unseen data.

Given the extreme class imbalance (~99% "No", ~1% "Yes"), we will rely on the **cost matrix** in Level 2 to handle the imbalance, rather than upsampling. This keeps the training distribution similar to the test distribution, which may improve generalization.

```{r}
set.seed(12345)

train_idx <- sample(1:nrow(br_filtered), 0.7 * nrow(br_filtered))
train_raw <- br_filtered[train_idx, ]
test <- br_filtered[-train_idx, ]

train_raw$finals <- factor(train_raw$finals, levels = c(0,1), labels = c("No", "Yes"))
test$finals <- factor(test$finals, levels = c(0,1), labels = c("No", "Yes"))

train_raw$cluster <- NULL
test$cluster <- NULL

train <- train_raw

# Get row indices in original data for test set
# First, find which rows in br_minmax correspond to br_filtered
filtered_indices <- which(br_minmax$cluster %in% c(1, 3))
# Then get test set indices
test_indices_in_filtered <- setdiff(1:nrow(br_filtered), train_idx)
test_indices_in_original <- filtered_indices[test_indices_in_filtered]

# Get metadata for test set from original data
test_metadata <- br_original[test_indices_in_original, ]

table(train$finals)
```

We use a **cost matrix** to handle class imbalance by adjusting the decision boundary. The cost matrix tells the model "predicting 'No' when it's actually 'Yes' (False Negative) is 10x worse than the reverse (False Positive)." This ensures the model prioritizes finding Final Four teams even with the extreme class imbalance, without needing to upsample the training data.

## 6 Build Level 1 Models

We'll build three separate models on the training data: Logistic Regression, Decision Tree, and Random Forest. KNN and SVM are excluded as they don't handle class imbalance well.

### 6.1 Logistic Regression

We start with a full logistic regression model.

```{r}
log_model <- glm(finals ~ ., data = train, family = binomial)
```

### 6.2 K-Nearest Neighbors

KNN is commented out as it doesn't handle class imbalance well.

```{r}
# knn_pred_raw <- knn(
#   train = train[, -which(names(train) == "finals")],
#   test  = test[, -which(names(test) == "finals")],
#   cl    = train$finals,
#   k     = 15
# )
```

### 6.3 Support Vector Machine (SVM)

SVM with RBF kernel is commented out as it doesn't handle class imbalance well.

```{r}
# svm_model <- ksvm(
#   finals ~ .,
#   data = train,
#   kernel = "rbfdot",
#   prob.model = TRUE
# )
```

### 6.4 Decision Tree (C5.0)

Next, we’ll build the standard C5.0 Decision Tree.

```{r}
dt_model <- C5.0(
  x = train[, -which(names(train) == "finals")],
  y = train$finals
)
```

### 6.5 Random Forest

We’ll add a Random Forest model, which creates multiple decision trees.

```{r}
rf_model <- randomForest(
  finals ~ .,
  data = train,
  ntree = 200,
  importance = TRUE
)
```

## 7 Level 1 Predictions and Evaluation

We now generate predictions on the **Test Set** for evaluation, and on the **Training Set** to prepare for Level 2 training.

```{r}
log_probs <- predict(log_model, test, type = "response")
log_pred <- ifelse(log_probs > 0.5, "Yes", "No")
log_pred <- factor(log_pred, levels = c("No", "Yes"))

# knn_pred_raw <- knn(
#   train = train[, -which(names(train) == "finals")],
#   test  = test[, -which(names(test) == "finals")],
#   cl    = train$finals,
#   k     = 15
# )

# svm_probs_matrix <- predict(svm_model, test, type = "probabilities")
# svm_probs <- svm_probs_matrix[, "Yes"] 
# svm_pred <- predict(svm_model, test)

dt_pred <- predict(dt_model, test)
dt_probs_matrix <- predict(dt_model, test, type = "prob")
dt_probs <- dt_probs_matrix[, "Yes"]

rf_pred <- predict(rf_model, test)
rf_probs_matrix <- predict(rf_model, test, type = "prob")
rf_probs <- rf_probs_matrix[, "Yes"]

cm_log <- confusionMatrix(log_pred, test$finals, positive = "Yes")
cm_log$table

# cm_knn <- confusionMatrix(knn_pred_raw, test$finals, positive = "Yes")
# cm_knn$table

# cm_svm <- confusionMatrix(svm_pred, test$finals, positive = "Yes")
# cm_svm$table

cm_dt  <- confusionMatrix(dt_pred, test$finals, positive = "Yes")
cm_dt$table

cm_rf  <- confusionMatrix(rf_pred, test$finals, positive = "Yes")
cm_rf$table

l1_results <- data.frame(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Kappa = c(cm_log$overall["Kappa"], cm_dt$overall["Kappa"], cm_rf$overall["Kappa"]),
  Accuracy = c(cm_log$overall["Accuracy"], cm_dt$overall["Accuracy"], cm_rf$overall["Accuracy"]),
  Sensitivity = c(cm_log$byClass["Sensitivity"], cm_dt$byClass["Sensitivity"], cm_rf$byClass["Sensitivity"])
)

kable(l1_results, digits = 3)

log_probs_train <- predict(log_model, train, type = "response")
# knn_pred_train <- knn(
#   train = train[, -which(names(train) == "finals")],
#   test  = train[, -which(names(train) == "finals")],
#   cl    = train$finals,
#   k     = 15
# )
# svm_probs_train <- predict(svm_model, train, type = "probabilities")[, "Yes"]
dt_probs_train <- predict(dt_model, train, type = "prob")[, "Yes"]
rf_probs_train <- predict(rf_model, train, type = "prob")[, "Yes"]
```

## 8 Build Level 2 Model (Stacking)

A single model, even a good one, has biases. We'll create a stacked model by feeding the predictions from the L1 models into a new meta-model (the L2 model).

We use the **Training Set** predictions to train this meta-model. Level 2 learns how to combine Level 1 predictions, and is evaluated on test set predictions to avoid data leakage.

```{r}
# knn_num_train <- ifelse(knn_pred_train == "Yes", 1, 0)

l2_train <- data.frame(
  LR_Prob = log_probs_train,
  DT_Prob = dt_probs_train,
  RF_Prob = rf_probs_train,
  finals = train$finals
)
# print(l2_train)

cost_matrix <- matrix(c(0, 700, 1, 0), nrow = 2, byrow = TRUE)
colnames(cost_matrix) <- c("No", "Yes")
rownames(cost_matrix) <- c("No", "Yes")

l2_model <- C5.0(
  x = l2_train[, -which(names(l2_train) == "finals")],
  y = l2_train$finals,
  costs = cost_matrix
)
```

## 9 Level 2 Predictions and Evaluation

Finally, we predict the Level 2 model on the **Test Set** (which it has never seen) and check performance. This allows for a fair apple-to-apple comparison with the Level 1 models evaluated in Section 7.

```{r}
# knn_num_test <- ifelse(knn_pred_raw == "Yes", 1, 0)

l2_test_input <- data.frame(
  LR_Prob = log_probs,
  DT_Prob = dt_probs,
  RF_Prob = rf_probs,
  finals = test$finals
)

l2_pred <- predict(l2_model, l2_test_input)
l2_cm <- confusionMatrix(l2_pred, test$finals, positive = "Yes")

l2_cm$table
```

## 9.1 Final Four Prediction Analysis

Now we analyze which Final Four teams were correctly predicted, incorrectly predicted, and missed. We add back the team names and seasons AFTER predictions are made to avoid any changes to the test data:

```{r}
# Create results dataframe with predictions and actual values
results_df <- data.frame(
  Season = test_metadata$Season,
  Team = test_metadata$Full.Team.Name,
  Predicted = l2_pred,
  Actual = test$finals
)

# Correctly predicted (True Positives)
correctly_predicted <- results_df[results_df$Predicted == "Yes" & results_df$Actual == "Yes", ]
cat("=== CORRECTLY PREDICTED FINAL FOUR TEAMS ===\n")
if(nrow(correctly_predicted) > 0) {
  print(correctly_predicted[, c("Season", "Team")])
} else {
  cat("None\n")
}

cat("\n=== INCORRECTLY PREDICTED FINAL FOUR TEAMS (False Positives) ===\n")
incorrectly_predicted <- results_df[results_df$Predicted == "Yes" & results_df$Actual == "No", ]
if(nrow(incorrectly_predicted) > 0) {
  print(incorrectly_predicted[, c("Season", "Team")])
} else {
  cat("None\n")
}

cat("\n=== MISSED FINAL FOUR TEAMS (False Negatives) ===\n")
missed <- results_df[results_df$Predicted == "No" & results_df$Actual == "Yes", ]
if(nrow(missed) > 0) {
  print(missed[, c("Season", "Team")])
} else {
  cat("None\n")
}
```

## 10 Comparison and Summary

First, we’ll look at the Kappas of all models.

```{r}
final_comparison <- rbind(
  l1_results,
  data.frame(Model = "Stacked L2 (Cost-Sensitive)", 
             Kappa = l2_cm$overall["Kappa"], 
             Accuracy = l2_cm$overall["Accuracy"],
             Sensitivity = l2_cm$byClass["Sensitivity"])
)

kable(final_comparison[order(-final_comparison$Kappa), ], digits = 3)
```

Next, we visualize the accuracy comparison.

```{r}
ggplot(final_comparison, aes(x = reorder(Model, Accuracy), y = Accuracy, fill = Model)) +
  geom_col() +
  coord_flip() +
  labs(title = "Model Accuracy Comparison", x = "", y = "Accuracy") +
  theme_minimal() +
  theme(legend.position = "none")
```

And importantly, since we care about finding "Yes" cases, we compare **Sensitivity**:

```{r}
ggplot(final_comparison, aes(x = reorder(Model, Sensitivity), y = Sensitivity, fill = Model)) +
  geom_col() +
  coord_flip() +
  labs(title = "Model Sensitivity Comparison", x = "", y = "Sensitivity (Recall)") +
  theme_minimal() +
  theme(legend.position = "none")
```