---
title: "GroupProject"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
date: "2025-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(caret)
library(class)
library(neuralnet)
library(kernlab)
library(C50)
library(reshape2)
library(dplyr) # for function mutate() in y_summary table
library(knitr) # for kable() in y_summary table
```

## Step 0 Why?

## Step 1 Load Data

```{r}
br <- read.csv("march.csv")

# Turn All String into Factor
br[sapply(br, is.character)] <- lapply(br[sapply(br, is.character)], as.factor)

# Rename Final.Four. into usable string "finals"
br <- br %>%
  rename(finals = Final.Four.)

br$finals <- ifelse(br$finals == "No", 0, 1)

# Make Summary Table to display No vs Yes quantities
summary <- br$Final.Four.
summary <- factor(summary, levels = c("No", "Yes"))
y_summary <- br |>
  group_by(finals) |>
  summarise(Count = n())|>
  mutate(Percent = round(Count / sum(Count) * 100, 1))

kable(y_summary)

```

## Step 2 Clean Data

We removed variations of redundant data columns such as Short Conference Name and Mapped Conference Name and Mapped ESPN Team Name. 
```{r}
names(Filter(is.factor, br))
# 1. Remove Unnecessary Factors
br <- br[, !sapply(br, is.factor) | (names(br) %in% c("finals", "Active.Coaching.Length"))]


# 2.remove rows before 2010, because much of the data was missing, dropped all rows in 2020 because there was no March Madness tournament that year
br <- subset(br, Season > 2010) 
br <- subset(br, Season != 2020)


# 3. Remove columns that have more than 1643 empty data rows
#br <- br[, colSums(is.na(br)) <= 1643]



# 4. Remove columns with zero variance (all 1s)
br$Post.Season.Tournament.Sorting.Index <- NULL

# Remove redundant cols
br$Since <- NULL
br$Active.Coaching.Length <- NULL


# Got rid of columns that have even 1 data row of  n/a
br <- br[, colSums(is.na(br)) == 0]


# Got rid of columns that have “Raw” in them
cols <- grepl("Raw", names(br))
br <- br[, !cols]


# Got rid of columns that have “Rank” in them
cols <- grepl("Rank", names(br))
br <- br[, !cols]


```

### Make Dummy Variables and Scale Data

```{r}
br_d <- as.data.frame(model.matrix(~ . -1, data = br))
minmax <- function(x){
  (x - min(x)) / (max(x) - min(x))
}

br_minmax <- as.data.frame(lapply(br_d, minmax))
```



## Step 3 Split Data

```{r}
trainfac <- 0.5
set.seed(12345)
trainrows <- sample(1:nrow(br_minmax), trainfac*nrow(br_minmax))

train <- br_minmax[trainrows, ]
test <- br_minmax[-trainrows, ]

str(br_minmax$finals)
train$finals <- factor(train$finals, levels = c(0,1))
test$finals <- factor(test$finals,  levels = c(0,1))
```


## Step 4 Build Clustering Model

```{r}
cluster_minmax <- as.data.frame(lapply(br_minmax, minmax))
#summary(cluster_minmax)

finals_backup <- br_minmax$finals

br_minmax$finals <- NULL

br_z <- as.data.frame(lapply(br_minmax, scale))
#summary(br_z)

set.seed(12345)
c6 <- kmeans(br_z, 5)
c6$size
c6$centers

br_z$cluster <- c6$cluster
br_z$finals <- finals_backup
```

```{r}
# Cluster Count / Percentage Horizontal Histogram
ggplot(br_z, aes(x = factor(cluster))) +
  geom_bar() +
  coord_flip() +
  geom_text(
    stat = "count",
    aes(label = paste0(after_stat(count), " (",
                   round(after_stat(count) / sum(after_stat(count)) * 100, 1),
                   "%)")),
    hjust = -0.1,
    size = 6,
    fontface = "bold"
  ) +
  labs(title = "Cluster Counts",
       x = "Cluster",
       y = "Count") +
expand_limits(y = max(table(br_z$cluster)) * 1.5)
```


## Step 5 Build First Level Models

### Logistic Regression

### K-Nearest Neighbors

TODO:
```{r}
knn_pred <- knn(
  train = train[, -which(names(train) == "finals")],
  test  = test[, -which(names(test) == "finals")],
  cl    = train$finals,
  k     = 103
)

cm_knn <- confusionMatrix(as.factor(knn_pred), as.factor(test$finals), positive = "1")
```


### Artificial Neural Network

### Support Vector Machine

### Decision Tree

### Randomn Forest

## Step 6 Build Second Level Model- Stacked Model




## Step 7 Model Evaluation

## Step 8 Conclusion