---
title: "HW#5: Predicting Tournament Success with Stacked Ensemble"
author: "Caleb Hyun"
date: "November 6, 2025"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(ggplot2)
library(caret)
library(class)
library(neuralnet)
library(kernlab)
library(C50)
library(reshape2)
library(dplyr)
library(knitr)
library(randomForest)
library(e1071)
library(pROC)
library(cluster) # for PAM

set.seed(12345)
```

## 1 Why do we Care?

Our goal is to identify which college basketball teams are most likely to make a Final Four appearance based on their regular season metrics. Early identification of these "Cinderella" or powerhouse teams allows athletic departments, analysts, and fans to allocate their attention and resources effectively. The "best" model will effectively identify tournament contenders (True Positives) while minimizing false hope for teams that won't make it deep (False Positives).

## 2 Load the Data

```{r}
br <- read.csv("march.csv")

# Turn All String into Factor
br[sapply(br, is.character)] <- lapply(br[sapply(br, is.character)], as.factor)

# Rename Final.Four. into usable string "finals"
br <- br %>%
  rename(finals = Final.Four.)
 
br$finals <- ifelse(br$finals == "No", 0, 1)

# Make Summary Table to display No vs Yes quantities
summary <- factor(br$finals, levels = c(0, 1), labels = c("No", "Yes"))
y_summary <- br |>
  group_by(finals) |>
  summarise(Count = n()) |>
  mutate(Percent = round(Count / sum(Count) * 100, 1))

kable(y_summary, col.names = c("Finals", "Count", "Percent"))
```

## 3 Clean The Data

Next, we clean the data. We remove variations of redundant data columns such as Short Conference Name and Mapped Conference Name and Mapped ESPN Team Name, as well as specific seasons that were problematic.

```{r}
# 1. Remove Unnecessary Factors
br <- br[, !sapply(br, is.factor) | (names(br) %in% c("finals", "Active.Coaching.Length"))]

# 2. Remove rows before 2010 and drop 2020
br <- subset(br, Season > 2010) 
br <- subset(br, Season != 2020)

# 3. Remove columns with zero variance (all 1s)
br$Post.Season.Tournament.Sorting.Index <- NULL

# Remove redundant cols
br$Since <- NULL
br$Active.Coaching.Length <- NULL

# Got rid of columns that have even 1 data row of n/a
br <- br[, colSums(is.na(br)) == 0]

# Got rid of columns that have “Raw” in them
cols <- grepl("Raw", names(br))
br <- br[, !cols]

# Got rid of columns that have “Rank” in them
cols <- grepl("Rank", names(br))
br <- br[, !cols]
```

Now we create dummy variables and scale the data. We use min-max scaling to normalize the feature range for distance-based algorithms like KNN.

```{r}
br_d <- as.data.frame(model.matrix(~ . -1, data = br))

minmax <- function(x){
  (x - min(x)) / (max(x) - min(x))
}

br_minmax <- as.data.frame(lapply(br_d, minmax))
```

## 4 Cluster Analysis

We use K-means clustering to segment the teams into 4 distinct groups, hoping to isolate different performance profiles. We then filter to keep only Clusters 1 and 3 for our predictive modeling.

```{r}
set.seed(12345)

# Temporarily remove target for clustering
br_cluster_data <- br_minmax[, names(br_minmax) != "finals"]

# K-means with k=4
k4 <- kmeans(br_cluster_data, centers = 4)

# Add cluster assignment back to data
br_minmax$cluster <- k4$cluster

# Filter to keep only Clusters 1 and 3
br_filtered <- subset(br_minmax, cluster %in% c(1, 3))
br_filtered$cluster <- factor(br_filtered$cluster) # Refactor

# Show counts for the kept clusters
table(br_filtered$cluster)
```

## 5 Split Train Test (Level 1)

We’ll split the **filtered data (Clusters 1 & 3)** into three sets:
1. **Training Set (50%)**: To train Level 1 models.
2. **Validation Set (35%)**: To train the Level 2 (Stacked) model.
3. **Test Set (15%)**: To evaluate ALL models fairly on unseen data.

Given the extreme class imbalance, we will **up-sample** the minority class in the training set only.

```{r}
set.seed(12345)

# 1. Split into Train (50%) and Rest (50%)
train_idx <- sample(1:nrow(br_filtered), 0.5 * nrow(br_filtered))
train_raw <- br_filtered[train_idx, ]
rest <- br_filtered[-train_idx, ]

# 2. Split Rest into Validation (70% of rest -> 35% total) and Test (30% of rest -> 15% total)
val_idx <- sample(1:nrow(rest), 0.7 * nrow(rest))
validation <- rest[val_idx, ]
test <- rest[-val_idx, ]

# Handle Class Imbalance: Up-Sampling on Train Only
train_raw$finals <- factor(train_raw$finals, levels = c(0,1), labels = c("No", "Yes"))
validation$finals <- factor(validation$finals, levels = c(0,1), labels = c("No", "Yes"))
test$finals <- factor(test$finals,  levels = c(0,1), labels = c("No", "Yes"))

# Drop cluster column from predictors (it was used for filtering)
train_raw$cluster <- NULL
validation$cluster <- NULL
test$cluster <- NULL

set.seed(12345)
train <- upSample(x = train_raw[, names(train_raw) != "finals"], 
                  y = train_raw$finals, 
                  yname = "finals")

# Check the new balance
table(train$finals)
```

We also **up-sample** the Validation Set. This is CRITICAL. If we train the Level 2 model on the original imbalanced validation set (where 99% are "No"), it will just learn to predict "No" for everything to maximize accuracy, ignoring the minority class signals from the Level 1 models.

```{r}
set.seed(12345)
validation_upsampled <- upSample(x = validation[, names(validation) != "finals"], 
                                 y = validation$finals, 
                                 yname = "finals")
```

## 6 Build Level 1 Models

We’ll build five separate models on the training data: Logistic Regression, K-Nearest Neighbors, Support Vector Machine, Decision Tree, and Random Forest.

### 6.1 Logistic Regression

We start with a full logistic regression model.

```{r}
log_model <- glm(finals ~ ., data = train, family = binomial)
```

### 6.2 K-Nearest Neighbors

For KNN, we will use a tuned k value.

```{r}
# Training KNN
knn_pred_raw <- knn(
  train = train[, -which(names(train) == "finals")],
  test  = test[, -which(names(test) == "finals")],
  cl    = train$finals,
  k     = 15
)
```

### 6.3 Support Vector Machine (SVM)

We will use a Radial Basis Function (RBF) kernel for the SVM.

```{r}
svm_model <- ksvm(
  finals ~ .,
  data = train,
  kernel = "rbfdot",
  prob.model = TRUE
)
```

### 6.4 Decision Tree (C5.0)

Next, we’ll build the standard C5.0 Decision Tree.

```{r}
dt_model <- C5.0(
  x = train[, -which(names(train) == "finals")],
  y = train$finals
)
```

### 6.5 Random Forest

We’ll add a Random Forest model, which creates multiple decision trees.

```{r}
rf_model <- randomForest(
  finals ~ .,
  data = train,
  ntree = 200,
  importance = TRUE
)
```

## 7 Level 1 Predictions and Evaluation

We now generate predictions on the **Test Set** for evaluation, and on the **Validation Set** to prepare for Level 2 training.

```{r}
# --- Predictions on Test Set (For Evaluation) ---
# 1. Logistic Regression
log_probs <- predict(log_model, test, type = "response")
log_pred <- ifelse(log_probs > 0.5, "Yes", "No")
log_pred <- factor(log_pred, levels = c("No", "Yes"))

# 2. KNN
knn_pred_raw <- knn(
  train = train[, -which(names(train) == "finals")],
  test  = test[, -which(names(test) == "finals")],
  cl    = train$finals,
  k     = 15
)

# 3. SVM
svm_probs_matrix <- predict(svm_model, test, type = "probabilities")
svm_probs <- svm_probs_matrix[, "Yes"] 
svm_pred <- predict(svm_model, test)

# 4. Decision Tree
dt_pred <- predict(dt_model, test)
dt_probs_matrix <- predict(dt_model, test, type = "prob")
dt_probs <- dt_probs_matrix[, "Yes"]

# 5. Random Forest
rf_pred <- predict(rf_model, test)
rf_probs_matrix <- predict(rf_model, test, type = "prob")
rf_probs <- rf_probs_matrix[, "Yes"]

# --- Evaluation on Test Set ---
cm_log <- confusionMatrix(log_pred, test$finals, positive = "Yes")
print("Logistic Regression CM:")
print(cm_log$table)

cm_knn <- confusionMatrix(knn_pred_raw, test$finals, positive = "Yes")
print("KNN CM:")
print(cm_knn$table)

cm_svm <- confusionMatrix(svm_pred, test$finals, positive = "Yes")
print("SVM CM:")
print(cm_svm$table)

cm_dt  <- confusionMatrix(dt_pred, test$finals, positive = "Yes")
print("Decision Tree CM:")
print(cm_dt$table)

cm_rf  <- confusionMatrix(rf_pred, test$finals, positive = "Yes")
print("Random Forest CM:")
print(cm_rf$table)

# Collect Results
l1_results <- data.frame(
  Model = c("Logistic Regression", "KNN", "SVM", "Decision Tree", "Random Forest"),
  Kappa = c(cm_log$overall["Kappa"], cm_knn$overall["Kappa"], cm_svm$overall["Kappa"], 
            cm_dt$overall["Kappa"], cm_rf$overall["Kappa"]),
  Accuracy = c(cm_log$overall["Accuracy"], cm_knn$overall["Accuracy"], cm_svm$overall["Accuracy"], 
               cm_dt$overall["Accuracy"], cm_rf$overall["Accuracy"]),
  Sensitivity = c(cm_log$byClass["Sensitivity"], cm_knn$byClass["Sensitivity"], cm_svm$byClass["Sensitivity"], 
                  cm_dt$byClass["Sensitivity"], cm_rf$byClass["Sensitivity"])
)

kable(l1_results, digits = 3)

# --- Predictions on Validation Set (For L2 Training) ---
# We predict on the UPSAMPLED validation set now
log_probs_val <- predict(log_model, validation_upsampled, type = "response")
knn_pred_val <- knn(
  train = train[, -which(names(train) == "finals")],
  test  = validation_upsampled[, -which(names(validation_upsampled) == "finals")],
  cl    = train$finals,
  k     = 15
)
svm_probs_val <- predict(svm_model, validation_upsampled, type = "probabilities")[, "Yes"]
dt_probs_val <- predict(dt_model, validation_upsampled, type = "prob")[, "Yes"]
rf_probs_val <- predict(rf_model, validation_upsampled, type = "prob")[, "Yes"]
```

## 8 Build Level 2 Model (Stacking)

A single model, even a good one, has biases. We’ll create a stacked model by feeding the predictions from the L1 models into a new meta-model (the L2 model).

We use the **Validation Set** predictions to train this meta-model, avoiding data leakage.

```{r}
# Prepare Level 2 Training Data (from Validation Set)
# Note: KNN from 'class' package only returns class labels, so we use 0/1 approximation if probs unavailable
knn_num_val <- ifelse(knn_pred_val == "Yes", 1, 0)

l2_train <- data.frame(
  LR_Prob = log_probs_val,
  KNN_Pred = knn_num_val, 
  SVM_Prob = svm_probs_val,
  DT_Prob = dt_probs_val,
  RF_Prob = rf_probs_val,
  finals = validation_upsampled$finals
)

# Build L2 Model (C5.0 with Cost Matrix)
# Penalize False Negatives (Missing a Final Four team) 50x more than False Positives
cost_matrix <- matrix(c(0, 1, 50, 0), nrow = 2, byrow = TRUE)
colnames(cost_matrix) <- c("No", "Yes")
rownames(cost_matrix) <- c("No", "Yes")

l2_model <- C5.0(
  x = l2_train[, -which(names(l2_train) == "finals")],
  y = l2_train$finals,
  costs = cost_matrix
)
```

## 9 Level 2 Predictions and Evaluation

Finally, we predict the Level 2 model on the **Test Set** (which it has never seen) and check performance. This allows for a fair apple-to-apple comparison with the Level 1 models evaluated in Section 7.

```{r}
# Prepare Test Data for L2 (from Test Set predictions generated in Section 7)
knn_num_test <- ifelse(knn_pred_raw == "Yes", 1, 0)

l2_test_input <- data.frame(
  LR_Prob = log_probs,
  KNN_Pred = knn_num_test,
  SVM_Prob = svm_probs,
  DT_Prob = dt_probs,
  RF_Prob = rf_probs,
  finals = test$finals
)

l2_pred <- predict(l2_model, l2_test_input)
l2_cm <- confusionMatrix(l2_pred, test$finals, positive = "Yes")

print("Stacked Model CM:")
print(l2_cm$table)
```

## 10 Comparison and Summary

First, we’ll look at the Kappas of all models.

```{r}
final_comparison <- rbind(
  l1_results,
  data.frame(Model = "Stacked L2 (Cost-Sensitive)", 
             Kappa = l2_cm$overall["Kappa"], 
             Accuracy = l2_cm$overall["Accuracy"],
             Sensitivity = l2_cm$byClass["Sensitivity"])
)

kable(final_comparison[order(-final_comparison$Kappa), ], digits = 3)
```

Next, we visualize the accuracy comparison.

```{r}
ggplot(final_comparison, aes(x = reorder(Model, Accuracy), y = Accuracy, fill = Model)) +
  geom_col() +
  coord_flip() +
  labs(title = "Model Accuracy Comparison", x = "", y = "Accuracy") +
  theme_minimal() +
  theme(legend.position = "none")
```

And importantly, since we care about finding "Yes" cases, we compare **Sensitivity**:

```{r}
ggplot(final_comparison, aes(x = reorder(Model, Sensitivity), y = Sensitivity, fill = Model)) +
  geom_col() +
  coord_flip() +
  labs(title = "Model Sensitivity Comparison", x = "", y = "Sensitivity (Recall)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 11 Summary and Discussion

### 11.1 Model Performance
We built and tuned five Level 1 models. The stacked model attempts to leverage the strengths of each. By using a cost-sensitive approach in Level 2, we prioritized finding the "Yes" cases (Final Four teams), even if it meant accepting slightly more False Positives.

## 12 Financial and Business Impact Analysis

### 12.1 The Business Problem
For an athletic director or a tournament betting syndicate, the cost of missing a potential Final Four team (False Negative) is often higher than betting on a team that fails (False Positive).

### 12.2 Cost-Benefit Framework
Let's assume:
- **Benefit (True Positive):** $50,000 (Revenue from deep tournament run / Winning bet payout)
- **Cost (False Positive):** -$5,000 (Wasted travel budget / Lost bet)
- **Cost (False Negative):** -$20,000 (Missed opportunity cost / Missed branding exposure)

```{r}
calculate_net_value <- function(cm) {
  TP <- cm$table[2, 2]
  FP <- cm$table[1, 2]
  FN <- cm$table[2, 1]
  TN <- cm$table[1, 1]
  
  benefit_TP <- 50000
  cost_FP <- 5000
  cost_FN <- 20000
  
  total_value <- (TP * benefit_TP) - (FP * cost_FP) - (FN * cost_FN)
  return(total_value)
}

# Calculate values for best L1 (Random Forest usually performs well) and L2
val_rf <- calculate_net_value(cm_rf)
val_l2 <- calculate_net_value(l2_cm)

impact_df <- data.frame(
  Model = c("Random Forest (L1)", "Stacked L2"),
  Net_Value = c(val_rf, val_l2)
)

kable(impact_df)
```

## 13 Part 2: Implementing the Prediction Model

### 13.1 Reverse Bias (Over-Ranking)
In college sports, teams often have inflated rankings due to "brand name" bias (e.g., Duke or Kentucky might be ranked higher than their stats justify). This is analogous to students lying about alcohol consumption—the "ground truth" (rankings/seedings) might be noisy.

**Impact:**
If the training labels (Final Four appearances) are truth, but our input features (Pre-tournament rankings) are biased, our model might learn to overvalue "blue blood" programs.

**Mitigation:**
To account for this, we rely more on "raw" stat columns (Points per Possession, Defensive Efficiency) rather than human-poll rankings (AP Poll) when possible, or build a separate model to detect "Overrated" teams by comparing their poll rank to their efficiency metrics.

### 13.2 Team Selection for Sponsorship
Suppose we need to sponsor 2 teams for the tournament, but we cannot afford a team that crashes out early. We want high confidence.

We would take the 5 finalists, run them through our Stacked L2 model, and look at the underlying probability scores. We would select the two teams with the highest predicted probability of "Yes" (Final Four), ensuring we maximize our brand exposure duration.

```{r}
# Example: Picking top 2 from the test set based on RF probability
test_w_probs <- data.frame(
  Actual = test$finals,
  RF_Prob = rf_probs,
  L2_Pred = ifelse(l2_pred == "Yes", 1, 0) 
)

# Sort by probability
top_picks <- head(test_w_probs[order(-test_w_probs$RF_Prob), ], 5)
kable(top_picks, caption = "Top 5 Teams by Predicted Probability")
```

**Recommendation:**
We select the top two teams from the probability list above. These teams show the strongest statistical profile consistent with historical Final Four contenders.
