---
title: "HW#5: Predicting Tournament Success with Stacked Ensemble"
author: "Caleb Hyun"
date: "November 6, 2025"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(ggplot2)
library(caret)
library(class)
library(neuralnet)
library(kernlab)
library(C50)
library(reshape2)
library(dplyr)
library(knitr)
library(randomForest)
library(e1071)
library(pROC)
library(cluster) # for PAM

set.seed(12345)
```

## 1 Why do we Care?

Our goal is to identify which college basketball teams are most likely to make a Final Four appearance based on their regular season metrics. Early identification of these "Cinderella" or powerhouse teams allows athletic departments, analysts, and fans to allocate their attention and resources effectively. The "best" model will effectively identify tournament contenders (True Positives) while minimizing false hope for teams that won't make it deep (False Positives).

## 2 Load the Data

```{r}
br <- read.csv("march.csv")

# Turn All String into Factor
br[sapply(br, is.character)] <- lapply(br[sapply(br, is.character)], as.factor)

# Rename Final.Four. into usable string "finals"
br <- br %>%
  rename(finals = Final.Four.)
 
br$finals <- ifelse(br$finals == "No", 0, 1)

# Make Summary Table to display No vs Yes quantities
summary <- factor(br$finals, levels = c(0, 1), labels = c("No", "Yes"))
y_summary <- br |>
  group_by(finals) |>
  summarise(Count = n()) |>
  mutate(Percent = round(Count / sum(Count) * 100, 1))

kable(y_summary, col.names = c("Finals", "Count", "Percent"))
```

## 3 Clean The Data

Next, we clean the data. We remove variations of redundant data columns such as Short Conference Name and Mapped Conference Name and Mapped ESPN Team Name, as well as specific seasons that were problematic.

```{r}
# 1. Remove Unnecessary Factors
br <- br[, !sapply(br, is.factor) | (names(br) %in% c("finals", "Active.Coaching.Length"))]

# 2. Remove rows before 2010 and drop 2020
br <- subset(br, Season > 2010) 
br <- subset(br, Season != 2020)

# 3. Remove columns with zero variance (all 1s)
br$Post.Season.Tournament.Sorting.Index <- NULL

# Remove redundant cols
br$Since <- NULL
br$Active.Coaching.Length <- NULL

# Got rid of columns that have even 1 data row of n/a
br <- br[, colSums(is.na(br)) == 0]

# Got rid of columns that have “Raw” in them
cols <- grepl("Raw", names(br))
br <- br[, !cols]

# Got rid of columns that have “Rank” in them
cols <- grepl("Rank", names(br))
br <- br[, !cols]
```

Now we create dummy variables and scale the data. We use min-max scaling to normalize the feature range for distance-based algorithms like KNN.

```{r}
br_d <- as.data.frame(model.matrix(~ . -1, data = br))

minmax <- function(x){
  (x - min(x)) / (max(x) - min(x))
}

br_minmax <- as.data.frame(lapply(br_d, minmax))
```

## 4 Cluster Analysis

We use K-means clustering to segment the teams into 4 distinct groups, hoping to isolate different performance profiles. We then filter to keep only Clusters 1 and 3 for our predictive modeling.

```{r}
set.seed(12345)

# Temporarily remove target for clustering
br_cluster_data <- br_minmax[, names(br_minmax) != "finals"]

# K-means with k=4
k4 <- kmeans(br_cluster_data, centers = 4)

# Add cluster assignment back to data
br_minmax$cluster <- k4$cluster

# Filter to keep only Clusters 1 and 3
br_filtered <- subset(br_minmax, cluster %in% c(1, 3))
br_filtered$cluster <- factor(br_filtered$cluster) # Refactor

# Show counts for the kept clusters
table(br_filtered$cluster)
```

## 5 Split Train Test (Level 1)

**Data Flow Summary:**
1. **Load and Clean Data**: Load raw data, clean columns, remove problematic seasons, create dummy variables, and apply min-max scaling.
2. **Cluster Analysis**: Perform K-means clustering (k=4) on the scaled data to identify distinct team performance profiles.
3. **Filter Clusters**: Keep only Clusters 1 and 3 for predictive modeling (removing Clusters 2 and 4).
4. **Split Data**: Divide the filtered data into:
   - **Training Set (50%)**: To train Level 1 models
   - **Validation Set (35%)**: To train the Level 2 (Stacked) model
   - **Test Set (15%)**: To evaluate ALL models fairly on unseen data
5. **Upsample Training Set**: Balance the training set by upsampling the minority class ("Yes" - Final Four teams).
6. **Upsample Validation Set**: Balance the validation set by upsampling the minority class. This ensures the Level 2 model has sufficient examples of both classes to learn from, even though we also use a cost matrix.
7. **Train Level 1 Models**: Build five models (Logistic Regression, KNN, SVM, Decision Tree, Random Forest) on the upsampled training set.
8. **Generate Level 1 Predictions**: 
   - Predict on **upsampled validation set** → used as features for Level 2 training
   - Predict on **original test set** → used for final evaluation
9. **Train Level 2 Model**: Build stacked model using Level 1 predictions from validation set, with cost matrix to penalize False Negatives.
10. **Evaluate**: Test Level 2 model on original test set predictions.

We'll split the **filtered data (Clusters 1 & 3)** into three sets:
1. **Training Set (50%)**: To train Level 1 models.
2. **Validation Set (35%)**: To train the Level 2 (Stacked) model.
3. **Test Set (15%)**: To evaluate ALL models fairly on unseen data.

Given the extreme class imbalance (~99% "No", ~1% "Yes"), we will **up-sample** the minority class in the training set. This ensures Level 1 models have sufficient examples of Final Four teams to learn meaningful patterns, rather than just defaulting to predicting "No" for everything.

```{r}
set.seed(12345)

# 1. Split into Train (50%) and Rest (50%)
train_idx <- sample(1:nrow(br_filtered), 0.5 * nrow(br_filtered))
train_raw <- br_filtered[train_idx, ]
rest <- br_filtered[-train_idx, ]

# 2. Split Rest into Validation (70% of rest -> 35% total) and Test (30% of rest -> 15% total)
val_idx <- sample(1:nrow(rest), 0.7 * nrow(rest))
validation <- rest[val_idx, ]
test <- rest[-val_idx, ]

# Handle Class Imbalance: Up-Sampling on Train Only
train_raw$finals <- factor(train_raw$finals, levels = c(0,1), labels = c("No", "Yes"))
validation$finals <- factor(validation$finals, levels = c(0,1), labels = c("No", "Yes"))
test$finals <- factor(test$finals,  levels = c(0,1), labels = c("No", "Yes"))

# Drop cluster column from predictors (it was used for filtering)
train_raw$cluster <- NULL
validation$cluster <- NULL
test$cluster <- NULL

set.seed(12345)
train <- upSample(x = train_raw[, names(train_raw) != "finals"], 
                  y = train_raw$finals, 
                  yname = "finals")

# Check the new balance
table(train$finals)
```

We also **up-sample** the Validation Set. We use **both upsampling and a cost matrix** because they address different aspects of class imbalance:

- **Upsampling** helps the model **LEARN** better patterns from the minority class by providing more examples. Without sufficient examples, the model may struggle to learn meaningful patterns even if it knows to prioritize the minority class.

- **Cost matrix** helps the model **MAKE DECISIONS** that prioritize the minority class by adjusting the decision boundary. It tells the model "predicting 'No' when it's actually 'Yes' is 200x worse than the reverse."

Together, upsampling ensures the model has learned good patterns from both classes, while the cost matrix ensures it applies those patterns in a way that prioritizes finding Final Four teams (even at the cost of more false positives).

```{r}
set.seed(12345)
validation_upsampled <- upSample(x = validation[, names(validation) != "finals"], 
                                 y = validation$finals, 
                                 yname = "finals")
```

## 6 Build Level 1 Models

We’ll build five separate models on the training data: Logistic Regression, K-Nearest Neighbors, Support Vector Machine, Decision Tree, and Random Forest.

### 6.1 Logistic Regression

We start with a full logistic regression model.

```{r}
log_model <- glm(finals ~ ., data = train, family = binomial)
```

### 6.2 K-Nearest Neighbors

For KNN, we will use a tuned k value.

```{r}
# Training KNN
knn_pred_raw <- knn(
  train = train[, -which(names(train) == "finals")],
  test  = test[, -which(names(test) == "finals")],
  cl    = train$finals,
  k     = 15
)
```

### 6.3 Support Vector Machine (SVM)

We will use a Radial Basis Function (RBF) kernel for the SVM.

```{r}
svm_model <- ksvm(
  finals ~ .,
  data = train,
  kernel = "rbfdot",
  prob.model = TRUE
)
```

### 6.4 Decision Tree (C5.0)

Next, we’ll build the standard C5.0 Decision Tree.

```{r}
dt_model <- C5.0(
  x = train[, -which(names(train) == "finals")],
  y = train$finals
)
```

### 6.5 Random Forest

We’ll add a Random Forest model, which creates multiple decision trees.

```{r}
rf_model <- randomForest(
  finals ~ .,
  data = train,
  ntree = 200,
  importance = TRUE
)
```

## 7 Level 1 Predictions and Evaluation

We now generate predictions on the **Test Set** for evaluation, and on the **Validation Set** to prepare for Level 2 training.

```{r}
# --- Predictions on Test Set (For Evaluation) ---
# 1. Logistic Regression
log_probs <- predict(log_model, test, type = "response")
log_pred <- ifelse(log_probs > 0.5, "Yes", "No")
log_pred <- factor(log_pred, levels = c("No", "Yes"))

# 2. KNN
knn_pred_raw <- knn(
  train = train[, -which(names(train) == "finals")],
  test  = test[, -which(names(test) == "finals")],
  cl    = train$finals,
  k     = 15
)

# 3. SVM
svm_probs_matrix <- predict(svm_model, test, type = "probabilities")
svm_probs <- svm_probs_matrix[, "Yes"] 
svm_pred <- predict(svm_model, test)

# 4. Decision Tree
dt_pred <- predict(dt_model, test)
dt_probs_matrix <- predict(dt_model, test, type = "prob")
dt_probs <- dt_probs_matrix[, "Yes"]

# 5. Random Forest
rf_pred <- predict(rf_model, test)
rf_probs_matrix <- predict(rf_model, test, type = "prob")
rf_probs <- rf_probs_matrix[, "Yes"]

# --- Evaluation on Test Set ---
cm_log <- confusionMatrix(log_pred, test$finals, positive = "Yes")
print("Logistic Regression CM:")
print(cm_log$table)

cm_knn <- confusionMatrix(knn_pred_raw, test$finals, positive = "Yes")
print("KNN CM:")
print(cm_knn$table)

cm_svm <- confusionMatrix(svm_pred, test$finals, positive = "Yes")
print("SVM CM:")
print(cm_svm$table)

cm_dt  <- confusionMatrix(dt_pred, test$finals, positive = "Yes")
print("Decision Tree CM:")
print(cm_dt$table)

cm_rf  <- confusionMatrix(rf_pred, test$finals, positive = "Yes")
print("Random Forest CM:")
print(cm_rf$table)

# Collect Results
l1_results <- data.frame(
  Model = c("Logistic Regression", "KNN", "SVM", "Decision Tree", "Random Forest"),
  Kappa = c(cm_log$overall["Kappa"], cm_knn$overall["Kappa"], cm_svm$overall["Kappa"], 
            cm_dt$overall["Kappa"], cm_rf$overall["Kappa"]),
  Accuracy = c(cm_log$overall["Accuracy"], cm_knn$overall["Accuracy"], cm_svm$overall["Accuracy"], 
               cm_dt$overall["Accuracy"], cm_rf$overall["Accuracy"]),
  Sensitivity = c(cm_log$byClass["Sensitivity"], cm_knn$byClass["Sensitivity"], cm_svm$byClass["Sensitivity"], 
                  cm_dt$byClass["Sensitivity"], cm_rf$byClass["Sensitivity"])
)

kable(l1_results, digits = 3)

# --- Predictions on Validation Set (For L2 Training) ---
# We predict on the UPSAMPLED validation set now
log_probs_val <- predict(log_model, validation_upsampled, type = "response")
knn_pred_val <- knn(
  train = train[, -which(names(train) == "finals")],
  test  = validation_upsampled[, -which(names(validation_upsampled) == "finals")],
  cl    = train$finals,
  k     = 15
)
svm_probs_val <- predict(svm_model, validation_upsampled, type = "probabilities")[, "Yes"]
dt_probs_val <- predict(dt_model, validation_upsampled, type = "prob")[, "Yes"]
rf_probs_val <- predict(rf_model, validation_upsampled, type = "prob")[, "Yes"]
```

## 8 Build Level 2 Model (Stacking)

A single model, even a good one, has biases. We’ll create a stacked model by feeding the predictions from the L1 models into a new meta-model (the L2 model).

We use the **Validation Set** predictions to train this meta-model, avoiding data leakage.

```{r}
# Prepare Level 2 Training Data (from Validation Set)
# Note: KNN from 'class' package only returns class labels, so we use 0/1 approximation if probs unavailable
knn_num_val <- ifelse(knn_pred_val == "Yes", 1, 0)

l2_train <- data.frame(
  LR_Prob = log_probs_val,
  KNN_Pred = knn_num_val, 
  SVM_Prob = svm_probs_val,
  DT_Prob = dt_probs_val,
  RF_Prob = rf_probs_val,
  finals = validation_upsampled$finals
)

# Build L2 Model (C5.0 with Cost Matrix)
# Penalize False Negatives (Missing a Final Four team) more than False Positives
# IMPORTANT: In C5.0, rows = PREDICTED classes, columns = ACTUAL classes
# Matrix format: [Predicted No, Actual No] [Predicted No, Actual Yes]
#                [Predicted Yes, Actual No] [Predicted Yes, Actual Yes]
# We want to penalize: Predicted No when Actual Yes (False Negative)
cost_matrix <- matrix(c(0, 5, 1, 0), nrow = 2, byrow = TRUE)
colnames(cost_matrix) <- c("No", "Yes")  # Actual classes
rownames(cost_matrix) <- c("No", "Yes")  # Predicted classes
# This means: False Negative (Predicted No, Actual Yes) costs 200
#             False Positive (Predicted Yes, Actual No) costs 1

l2_model <- C5.0(
  x = l2_train[, -which(names(l2_train) == "finals")],
  y = l2_train$finals,
  costs = cost_matrix
)
```

## 9 Level 2 Predictions and Evaluation

Finally, we predict the Level 2 model on the **Test Set** (which it has never seen) and check performance. This allows for a fair apple-to-apple comparison with the Level 1 models evaluated in Section 7.

```{r}
# Prepare Test Data for L2 (from Test Set predictions generated in Section 7)
knn_num_test <- ifelse(knn_pred_raw == "Yes", 1, 0)

l2_test_input <- data.frame(
  LR_Prob = log_probs,
  KNN_Pred = knn_num_test,
  SVM_Prob = svm_probs,
  DT_Prob = dt_probs,
  RF_Prob = rf_probs,
  finals = test$finals
)

l2_pred <- predict(l2_model, l2_test_input)
l2_cm <- confusionMatrix(l2_pred, test$finals, positive = "Yes")

print("Stacked Model CM:")
print(l2_cm$table)
```

## 10 Comparison and Summary

First, we’ll look at the Kappas of all models.

```{r}
final_comparison <- rbind(
  l1_results,
  data.frame(Model = "Stacked L2 (Cost-Sensitive)", 
             Kappa = l2_cm$overall["Kappa"], 
             Accuracy = l2_cm$overall["Accuracy"],
             Sensitivity = l2_cm$byClass["Sensitivity"])
)

kable(final_comparison[order(-final_comparison$Kappa), ], digits = 3)
```

Next, we visualize the accuracy comparison.

```{r}
ggplot(final_comparison, aes(x = reorder(Model, Accuracy), y = Accuracy, fill = Model)) +
  geom_col() +
  coord_flip() +
  labs(title = "Model Accuracy Comparison", x = "", y = "Accuracy") +
  theme_minimal() +
  theme(legend.position = "none")
```

And importantly, since we care about finding "Yes" cases, we compare **Sensitivity**:

```{r}
ggplot(final_comparison, aes(x = reorder(Model, Sensitivity), y = Sensitivity, fill = Model)) +
  geom_col() +
  coord_flip() +
  labs(title = "Model Sensitivity Comparison", x = "", y = "Sensitivity (Recall)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 11 Summary and Discussion

## 12 Financial and Business Impact Analysis

## 13 Part 2: Implementing the Prediction Model

### 13.1 Reverse Bias (Over-Ranking)

### 13.2 Team Selection for Sponsorship
